% Zero-shot texture-segmentation comparison — narrative discussion.
% Generated manually after running master_runner.py on all four datasets.
% Numbers match Tables~\ref{tab:miou} and \ref{tab:ari}, produced by
% aggregate_results.py (summary.tex / summary_ari.tex).
%
% Include into the main manuscript with:  \input{results/discussion.tex}
%
\subsection{Zero-Shot Comparison}
\label{sec:zs_comparison}

We evaluate four families that can turn a natural-language query into a
per-region segmentation mask without any task-specific training:
\texttt{sam3\_vanilla} grounds short class phrases through SAM3's native
semantic-segmentation head and switches to SAM3's mask-proposal decoder on
ADE20K, where no class vocabulary is available;
\texttt{grounded\_sam3} runs Grounding\,DINO on a per-class phrase list
and feeds the top-$K$ boxes into SAM3's geometry encoder;
\texttt{sa2va} is the end-to-end vision-language baseline (ByteDance
Sa2VA-4B) prompted with the interleaved-segmentation caption format it was
trained on;
and \texttt{qwen2sam\_zs} is our proposed Qwen3-VL\,+\,SAM3 pipeline.
Every family is scored with the identical post-processing stack—softmax
against a shared dustbin logit, Hungarian matching between predicted and
ground-truth masks, and a uniform $-10^{4}$ logit pad applied whenever a
baseline returns fewer than $K_\text{GT}$ regions—so that under-producing
masks incurs a bounded penalty rather than being silently rewarded. All
runs use $n = 3091$ (CAID), $253$ (RWTD), $200$ (STLD), and $212$
(ADE20K-Detecture) test samples respectively.

The proposed Qwen2SAM\,ZS pipeline achieves the strongest mean across
every metric we report: $0.834$ matched-mean IoU, $0.876$ Dice, and
$0.798$ adjusted Rand index, leading the runner-up (Sa2VA-4B) by
$+11.8$, $+11.8$, and $+9.1$ percentage points respectively on the
average over the four datasets. The margin is largest on the synthetic
benchmarks where class identity must be inferred from texture alone
(STLD, $+14.0$ mIoU over the best baseline) and on ADE20K-Detecture,
where the number of regions varies per image and the VLM component of
our pipeline has to discover $K$ itself ($+6.4$ mIoU over Sa2VA). On
real-world texture pairs (RWTD), the advantage persists but shrinks
($+8.0$ mIoU over Grounding-DINO\,+\,SAM3), suggesting that box-level
priors close part of the gap when texture regions happen to coincide
with salient foreground objects.

The single dataset where a simpler baseline outperforms our pipeline is
CAID, on which \texttt{sam3\_vanilla} reaches $0.901$ mIoU and $0.817$
ARI against our $0.866$ and $0.773$. The gap is instructive rather than
damaging: CAID poses a single, linguistically unambiguous class
(``water surface''), and feeding that exact phrase directly into SAM3's
text encoder is the shortest path from prompt to mask. Any additional
reasoning step---Qwen describing the scene, Grounding-DINO drawing a
box, or Sa2VA emitting a paragraph with interleaved \texttt{[SEG]}
tokens---adds noise without adding signal. Our pipeline's $0.773$ ARI
on the same dataset is nonetheless above every other baseline except
\texttt{sam3\_vanilla}, indicating that when the class is already
canonical, elaboration costs a few IoU points but still produces a
consistent partition. Practically, CAID argues for a fast-path that
skips the VLM when the user supplies a clean class term; the rest of
our results demonstrate why the VLM is needed whenever they do not.

The mIoU--ARI relationship is the most useful diagnostic across
baselines. \texttt{grounded\_sam3} exhibits the widest mIoU--ARI
dispersion on CAID ($0.622$ vs. $0.404$), where irregular shorelines
expose the limits of axis-aligned box priors: the Hungarian matcher
can still pair a coarsely correct box with a GT mask, but the pixel-
wise partition leaks across the waterline. Sa2VA shows the opposite
imbalance on ADE20K-Detecture: its mIoU ($0.631$) is modest relative to
its ARI ($0.698$), indicating that its interleaved \texttt{[SEG]}
masks describe the scene coherently even when any individual overlap
with GT is imperfect. Our method keeps the two metrics close on every
dataset (mIoU--ARI gap of at most $0.09$ where mIoU leads, and an
\emph{inversion}---ARI exceeding mIoU---on ADE20K-Detecture), a direct
consequence of routing the VLM's per-region grounding through SAM3
rather than generating masks from scratch: grounding preserves boundary fidelity, while the
language-side caption preserves partition coherence. The combined
mIoU and ARI columns of Table~\ref{tab:miou} thus support the same
conclusion: a VLM\,+\,grounded-segmenter composition outperforms
either component in isolation, whether that component is a pure text
encoder (\texttt{sam3\_vanilla}), a detection-prior box decoder
(\texttt{grounded\_sam3}), or an end-to-end conversational segmenter
(\texttt{sa2va}).
